---
title: "Introduction to DGVMTools"
author: "Matthew Forrest"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

DGVMTools is designed to provide core functionality for developing and utilising DGVMs (Dynamic Global Vegetation Models) and similar models (such as Land Surface Models). The basic motivation is that many people work with such models but their is no centralised effort to provide convenient, reliable and efficicent tools to perform common tasks such as reading model data, aggregating it in space, time and across vegetation types, producing plots, comparing to standard observed datasets etc.  Instead, people develop (and possibly share) scripts on an _ad hoc_ basis, often resulting in duplicated effort and technical limitations hindering scientific analysis.  

In fact much of the package could be utilised for any data which follows comes distributed over space (longitude, latitude), time (multiple years of data with subannual time resolution of daily, monthly or annual) and conceptual layers (for example different vegetation types or soil carbon pools).  However it has several keystone features which are focussed on vegetation modelling, namely: the ability to read data from selected DGVMs in their native formats (currently LPJ-GUESS(-SPITFIRE) and aDGVM2 are supported); facilities for handling metadata about Plant Functional Types (PFTs); and the means to classify vegetation model output into biomes.  





## BLAH

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
